---
title: "Introduction to R"
author: "Data Carpentry contributors"
minutes: 45
layout: topic
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

------------

> ## Learning Objectives
>
> * Assign names to objects in R with <- and =.
> * Solve mathematical operations in R.
> * Describe what a function is in R.
> * Describe what vectors are and how they can be manipulated in R.
> * Inspect the content of vectors in R and describe their content with class and str.

------------

## The R syntax

_Start by showing an example of a script_

* Point to the different parts:
  - a function
  - the assignment operator `<-`
  - the `=` for arguments
  - the comments `#` and how they are used to document function and its content
  - the `$` operator
* Point to indentation and consistency in spacing to improve clarity


## Creating objects

```{r, echo=FALSE, purl=TRUE}
### Creating objects (assignments)
```

You can get an output from R simply by typing in math in the console

```{r, purl=FALSE}
3 + 5
12/7
```

We can also comment on what it is that we're doing

```{r, purl=FALSE}
# I'm adding 3 and 5. R is fun!
3+5
```

What happens if we type that same command without the # sign in the front?

```{r, purl=FALSE, eval=FALSE}
I'm adding 3 and 5. R is fun!
3+5
```

Now R is trying to run that sentence as a command, and it
doesn't work. Now we're stuck over in the console. The
`+` sign means that it's still waiting for input, so we
can't type in a new command. To get out of this press the <kbd>Esc</kbd> key. This will work whenever you're stuck with that `+` sign.

It's great that R is a glorified caluculator, but obviously
we want to do more interesting things.

To do useful and interesting things, we need to assign _values_ to
_objects_. To create objects, we need to give it a name followed by the
assignment operator `<-` and the value we want to give it.

For instance, instead of adding 3 + 5, we can assign those
values to objects and then add them.

```{r, purl=FALSE}
# assign 3 to a
a <- 3
# assign 5 to b
b <- 5

# what now is a
a
# what now is b
b

#Add a and b
a + b
```

`<-` is the assignment operator. It assigns values on the right to objects on
the left. So, after executing `x <- 3`, the value of `x` is `3`. The arrow can
be read as 3 **goes into** `x`.  You can also use `=` or `->`for assignments but not in
all contexts so it is good practice to use `<-` for assignments. `=` should only
be used to specify the values of arguments in functions, see below.

In RStudio, typing <kbd>Alt</kbd> + <kbd>-</kbd> (push <kbd>Alt</kbd> at the same time as the <kbd>-</kbd> key) will write ` <- ` in a single keystroke.

### Exercise

- What happens if we change `a` and then re-add `a` and `b`?
- Does it work if you just change `a` in the script and then add `a` and `b`? Did you still get the same answer after they changed `a`? If so, why do you think that might be?
- We can also assign a + b to a new variable, `c`. How would you do this?

## Notes on objects

Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. You want your object names to be explicit and not too long. They
cannot start with a number (`2x` is not valid but `x2` is). R is case sensitive
(e.g., `Genome_length_mb` is different from `genome_length_mb`). There are some names that
cannot be used because they represent the names of fundamental functions in R
(e.g., `if`, `else`, `for`, see
[here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
for a complete list). In general, even if it's allowed, it's best to not use
other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`). When in
doubt, check the help to see if the name is already in use. It's also best to avoid
dots (`.`) within a variable name as in `my.dataset`. There are many functions
in R with dots in their names for historical reasons, but because dots have a
special meaning in R (for methods) and other programming languages, it's best to
avoid them. It is also recommended to use nouns for variable names, and verbs
for function names. It's important to be consistent in the styling of your code
(where you put spaces, how you name variables, etc.). In R, two popular style
guides are [Hadley Wickham's](http://adv-r.had.co.nz/Style.html) and
[Google's](https://google.github.io/styleguide/Rguide.xml).

When assigning a value to an object, R does not print anything. You can force to
print the value by using parentheses or by typing the name:

```{r}
# Assigns a value to a variable
genome_size_mb <- 35

# Assigns a value to a variable and prints it out on the console
(genome_size_mb <- 35)

# Prints out the value of a variable on the console
genome_size_mb
```

## Functions

The other key feature of R are functions. These are R's built in capabilities. Some examples of these are mathematical functions, like
`sqrt` and `round`. You can also get functions from libraries (which we'll talk about in a bit), or even write your own.

Functions are "canned scripts" that automate something complicated or convenient
or both.  Many functions are predefined, or become available when using the
function `library()` (more on that later). A function usually gets one or more
inputs called *arguments*. Functions often (but not always) return a *value*. A
typical example would be the function `sqrt()`. The input (the argument) must be
a number, and the return value (in fact, the output) is the square root of that
number. Executing a function ('running it') is called *calling* the function. An
example of a function call is:

`sqrt(a)`

Here, the value of `a` is given to the `sqrt()` function, the `sqrt()` function
calculates the square root. This function is very simple, because it takes just one
argument.

The return 'value' of a function need not be numerical (like that of `sqrt()`),
and it also does not need to be a single item: it can be a set of things, or
even a data set. We'll see that when we read data files in to R.

Arguments can be anything, not only numbers or filenames, but also other
objects. Exactly what each argument means differs per function, and must be
looked up in the documentation (see below). If an argument alters the way the
function operates, such as whether to ignore 'bad values', such an argument is
sometimes called an *option*.

Most functions can take several arguments, but many have so-called *defaults*.
If you don't specify such an argument when calling the function, the function
itself will fall back on using the *default*. This is a standard value that the
author of the function specified as being "good enough in standard cases". An
example would be what symbol to use in a plot. However, if you want something
specific, simply change the argument yourself with a value of your choice.

Let's try a function that can take multiple arguments `round`.

```{r, results='show'}
round(3.14159)
```

We can see that we get `3`. That's because the default is to round
to the nearest whole number. If we want more digits we can see
how to do that by getting information about the `round` function.
We can use `args(round)` or look at the
help for this function using `?round`.

```{r, results='show'}
args(round)
```

```{r, eval=FALSE}
?round
```


We see that if we want a different number of digits, we can
type `digits=2` or however many we want.

```{r, results='show'}
round(3.14159, digits=2)
```

If you provide the arguments in the exact same order as they are defined you don't have to name them:

```{r, results='show'}
round(3.14159, 2)
```

However, it's usually not recommended practice because it's a lot of remembering
to do, and if you share your code with others that includes less known functions
it makes your code difficult to read. (It's however OK to not include the names
of the arguments for basic functions like `mean`, `min`, etc...)

Another advantage of naming arguments, is that the order doesn't matter.
This is useful when there start to be more arguments.


### Exercise

We're going to work with genome lengths.
- Create a variable `genome_length_mb` and assign it the value `4.6`

#### Solution

```{r, purl=FALSE}
genome_length_mb <- 4.6
genome_length_mb
```

### Exercise

Now that R has `genome_length_mb` in memory, we can do arithmetic with it. For
instance, we may want to convert this to the weight of the genome in
picograms (for some reason). 978Mb = 1picogram. Divide the
genome length in Mb by 978.

#### Solution

```{r, purl=FALSE}
genome_length_mb / 978.0
```

It turns out an E. coli genome doesn't weigh very much.

### Exercise

We can also change the variable's value by assigning it a new one. Say
we want to think about a human genome rather than E. coli. Change `genome_length_mb` to 3000 and figure out the weight of the human
genome.

#### Solution

```{r, purl=FALSE}
genome_length_mb <- 3000.0
genome_length_mb / 978.0
```

This means that assigning a value to one variable does not change the values of
other variables.  For example, let's store the genome's weight in a variable.

```{r, purl=FALSE}
genome_weight_pg <- genome_length_mb / 978.0
```

and then change `genome_length_mb` to 100.

```{r, purl=FALSE}
genome_length_mb <- 100
```

### Exercise

What do you think is the current content of the object `genome_weight_pg`? 3.06 or 0.102?


## Vectors and data types

```{r, echo=FALSE, purl=TRUE}
### Vectors and data types
```

A vector is the most common and basic data structure in R, and is pretty much
the workhorse of R. It's basically just a list of values, mainly either numbers or
characters. They're special lists that you can do math with. You can assign this list of values to a variable, just like you
would for one item. For example we can create a vector of genome lengths:

```{r, purl=FALSE}
glengths <- c(4.6, 3000, 50000)
glengths
```

A vector can also contain characters:

```{r, purl=FALSE}
species <- c("ecoli", "human", "corn")
species
```

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r, purl=FALSE}
length(glengths)
length(species)
```


You can also do math with whole vectors. For instance if we wanted to multiply the genome lengths of all the genomes in the list, we can do

```{r, purl=FALSE}
5 * glengths
```

or we can add the data in the two vectors together

```{r, purl=FALSE}
new_lengths <- glengths + glengths
new_lengths
```

This is very useful if we have data in different vectors that we
want to combine or work with.

There are few ways to figure out what's going on in a vector.

`class()` indicates the class (the type of element) of an object:

```{r, purl=FALSE}
class(glengths)
class(species)
```

The function `str()` provides an overview of the object and the elements it
contains. It is a really useful function when working with large and complex
objects:

```{r, purl=FALSE}
str(glengths)
str(species)
```

You can add elements to your vector simply by using the `c()` function:

```{r, purl=FALSE}
lengths <- c(glengths, 90) # adding at the end
lengths <- c(30, glengths) # adding at the beginning
lengths
```

What happens here is that we take the original vector `glengths`, and we are
adding another item first to the end of the other ones, and then another item at
the beginning. We can do this over and over again to build a vector or a
dataset. As we program, this may be useful to autoupdate results that we are
collecting or calculating.

We just saw 2 of the 6 **data types** that R uses: `"character"` and
`"numeric"`. The other 4 are:       

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
  `1+4i`) and that's all we're going to say about them
* `"raw"` that we won't discuss further

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`) and
factors (`factor`).


## Seeking help

### I know the name of the function I want to use, but I'm not sure how to use it

If you need help with a specific function, let's say `barplot()`, you can type:

```{r, eval=FALSE}
?barplot
```

If you just need to remind yourself of the names of the arguments, you can use:

```{r, eval=FALSE}
args(lm)
```

If the function is part of a package that is installed on your computer but
don't remember which one, you can type:

```{r, eval=FALSE}
??geom_point
```

### I want to use a function that does X, there must be a function for it but I don't know which one...

If you are looking for a function to do a particular task, you can use
`help.search()` (but only looks through the installed packages):

```{r, eval=FALSE}
help.search("kruskal")
```

If you can't find what you are looking for, you can use the
[rdocumention.org](http://www.rdocumentation.org) website that search through
the help files across all packages available.

### I am stuck... I get an error message that I don't understand

Start by googling the error message. However, this doesn't always work very well
because often, package developers rely on the error catching provided by R. You
end up with general error messages that might not be very helpful to diagnose a
problem (e.g. "subscript out of bounds").

However, you should check [stackoverflow.com](https://stackoverflow.com). Search
using the `[r]` tag. Most questions have already been answered, but the
challenge is to use the right words in the search to find the answers:
[http://stackoverflow.com/questions/tagged/r](http://stackoverflow.com/questions/tagged/r)

The [Introduction to R](http://cran.r-project.org/doc/manuals/R-intro.pdf) can
also be dense for people with little programming experience but it is a good
place to understand the underpinnings of the R language.

The [R FAQ](http://cran.r-project.org/doc/FAQ/R-FAQ.html) is dense and technical
but it is full of useful information.

### Asking for help

The key to get help from someone is for them to grasp your problem rapidly. You
should make it as easy as possible to pinpoint where the issue might be.

Try to use the correct words to describe your problem. For instance, a package
is not the same thing as a library. Most people will understand what you meant,
but others have really strong feelings about the difference in meaning. The key
point is that it can make things confusing for people trying to help you. Be as
precise as possible when describing your problem

If possible, try to reduce what doesn't work to a simple reproducible
example. If you can reproduce the problem using a very small `data.frame`
instead of your 50,000 rows and 10,000 columns one, provide the small one with
the description of your problem. When appropriate, try to generalize what you
are doing so even people who are not in your field can understand the question.

To share an object with someone else, if it's relatively small, you can use the
function `dput()`. It will output R code that can be used to recreate the exact same
object as the one in memory:

```{r, results='show'}
dput(head(iris)) # iris is an example data.frame that comes with R
```

If the object is larger, provide either the raw file (i.e., your CSV file) with
your script up to the point of the error (and after removing everything that is
not relevant to your issue). Alternatively, in particular if your questions is
not related to a `data.frame`, you can save any R object to a file. Note: for this
example, the folder "/tmp" needs to already exist.

```{r, eval=FALSE}
saveRDS(iris, file="/tmp/iris.rds")
```

The content of this file is however not human readable and cannot be posted
directly on stackoverflow. It can however be sent to someone by email who can read
it with this command:

```{r, eval=FALSE}
some_data <- readRDS(file="~/Downloads/iris.rds")
```

Last, but certainly not least, **always include the output of `sessionInfo()`**
as it provides critical information about your platform, the versions of R and
the packages that you are using, and other information that can be very helpful
to understand your problem.

```{r, results='show'}
sessionInfo()
```

### Where to ask for help?

* Your friendly colleagues: if you know someone with more experience than you,
  they might be able and willing to help you.
* Stackoverflow: if your question hasn't been answered before and is well
  crafted, chances are you will get an answer in less than 5 min.
* The [R-help](https://stat.ethz.ch/mailman/listinfo/r-help): it is read by a
  lot of people (including most of the R core team), a lot of people post to it,
  but the tone can be pretty dry, and it is not always very welcoming to new
  users. If your question is valid, you are likely to get an answer very fast
  but don't expect that it will come with smiley faces. Also, here more than
  everywhere else, be sure to use correct vocabulary (otherwise you might get an
  answer pointing to the misuse of your words rather than answering your
  question). You will also have more success if your question is about a base
  function rather than a specific package.
* If your question is about a specific package, see if there is a mailing list
  for it. Usually it's included in the DESCRIPTION file of the package that can
  be accessed using `packageDescription("name-of-package")`. You may also want
  to try to email the author of the package directly.
* There are also some topic-specific mailing lists (GIS, phylogenetics, etc...),
  the complete list is [here](http://www.r-project.org/mail.html).

### More resources

* The [Posting Guide](http://www.r-project.org/posting-guide.html) for the R
  mailing lists.
* [How to ask for R help](http://blog.revolutionanalytics.com/2014/01/how-to-ask-for-r-help.html)
  useful guidelines

