---
title: Introduction to R
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/')
```

------------

> ## Learning Objectives
>
> * Familiarize participants with R syntax
> * Understand the concepts of objects and assignment
> * Understand the concepts of vector and data types
> * Get exposed to a few functions

------------

## The R syntax


## Creating objects

```{r, echo=FALSE, purl=TRUE}
### Creating objects (assignments)
```

You can get output from R simply by typing in math in the console

```{r, purl=FALSE}
3 + 5
12/7
2*4
2^4
```

We can also comment what it is we're doing

```{r, purl=FALSE}
# I am adding 3 and 5. R is fun!
3 + 5
```

What happens if we do that same command with the # sign in the front?

```{r, purl=FALSE, eval=FALSE}
I am adding 3 and 5. R is fun!
3 + 5
```

Now R is trying to run that sentence as a command, and it 
doesn't work. Now we're stuck over in the console. The 
`+` sign means that it's still waiting for input, so we 
can't type in a new command. To get out of this type `Esc`. This will work whenever you're stuck with that `+` sign. 

It's great that R is a glorified caluculator, but obviously
we want to do more interesting things. 

To do useful and interesting things, we need to assign _values_ to
_objects_. To create objects, we need to give it a name followed by the
assignment operator `<-` and the value we want to give it.

## Assignment operator


For instance, instead of adding 3 + 5, we can assign those
values to objects and then add them.

```{r, purl=FALSE}
# assign 3 to a
a <- 3
# assign 5 to b
b <- 5

# what now is a
a
# what now is b
b

#Add a and b
a + b
```

`<-` is the assignment operator. It assigns values on the right to objects on
the left. So, after executing `x <- 3`, the value of `x` is `3`. The arrow can
be read as 3 **goes into** `x`.  You can also use `=` for assignments but not in
all contexts so it is good practice to use `<-` for assignments. `=` should only
be used to specify the values of arguments in functions, see below.

In RStudio, typing `Alt + -` (push `Alt`, the key next to your space bar at the
same time as the `-` key) will write ` <- ` in a single keystroke.

To view which objects we have stored in memory, we can use the `ls()` command

```{r, purl=FALSE}
ls()
```

To remove objects we can use the `rm()` command
```{r, purl=FALSE}
rm(a)
```
### Exercise 

- What happens if we change `a` and then re-add `a` and `b`?
- Does it work if you just change `a` in the script and then add `a` and `b`? Did you still get the same answer after they changed `a`? If so, why do you think that might be?
- We can also assign a + b to a new variable, `c`. How would you do this?

## Notes on objects

Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. You want your object names to be explicit and not too long. They
cannot start with a number (`2x` is not valid but `x2` is). R is case sensitive
(e.g., `my_data` is different from `My_data`). There are some names that
cannot be used because they represent the names of fundamental functions in R
(e.g., `if`, `else`, `for`, see
[here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
for a complete list). In general, even if it's allowed, it's best to not use
other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`). In doubt
check the help to see if the name is already in use. It's also best to avoid
dots (`.`) within a variable name as in `my.dataset`. There are many functions
in R with dots in their names for historical reasons, but because dots have a
special meaning in R (for methods) and other programming languages, it's best to
avoid them. It is also recommended to use nouns for variable names, and verbs
for function names. It's important to be consistent in the styling of your code
(where you put spaces, how you name variable, etc.). In R, two popular style
guides are [Hadley Wickham's](http://adv-r.had.co.nz/Style.html) and
[Google's](https://google-styleguide.googlecode.com/svn/trunk/Rguide.xml).

When assigning a value to an object, R does not print anything. You can force to
print the value by using parentheses or by typing the name:


## Functions

The other key feature of R are functions. These are R's built in capabilities. Some examples of these are mathematical functions, like
`sqrt` and `round`. You can also get functions from libraries (which we'll talk about in a bit), or even write your own. 

## Functions and their arguments

Functions are "canned scripts" that automate something complicated or convenient
or both.  Many functions are predefined, or become available when using the
function `library()` (more on that later). A function usually gets one or more
inputs called *arguments*. Functions often (but not always) return a *value*. A
typical example would be the function `sqrt()`. The input (the argument) must be
a number, and the return value (in fact, the output) is the square root of that
number. Executing a function ('running it') is called *calling* the function. An
example of a function call is:

`sqrt(a)`

Here, the value of `a` is given to the `sqrt()` function, the `sqrt()` function
calculates the square root. This function is very simple, because it takes just one
argument. 

The return 'value' of a function need not be numerical (like that of `sqrt()`),
and it also does not need to be a single item: it can be a set of things, or
even a data set. We'll see that when we read data files in to R. 

Arguments can be anything, not only numbers or filenames, but also other
objects. Exactly what each argument means differs per function, and must be
looked up in the documentation (see below). If an argument alters the way the
function operates, such as whether to ignore 'bad values', such an argument is
sometimes called an *option*.

Most functions can take several arguments, but many have so-called *defaults*.
If you don't specify such an argument when calling the function, the function
itself will fall back on using the *default*. This is a standard value that the
author of the function specified as being "good enough in standard cases". An
example would be what symbol to use in a plot. However, if you want something
specific, simply change the argument yourself with a value of your choice.

Let's try a function that can take multiple arguments `round`.

```{r, results='show'}
round(3.14159)
```

We can see that we get `3`. That's because the default is to round
to the nearest whole number. If we want more digits we can see 
how to do that by getting information about the `round` function.
We can use `args(round)` or look at the
help for this function using `?round`.

```{r, results='show'}
args(round)
```

```{r, eval=FALSE}
?round
```


We see that if we want a different number of digits, we can 
type `digits=2` or however many we want.

```{r, results='show'}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined you don't have to name them:

```{r, results='show'}
round(3.14159, 2)
```

However, it's usually not recommended practice because it's a lot of remembering
to do, and if you share your code with others that includes less known functions
it makes your code difficult to read. (It's however OK to not include the names
of the arguments for basic functions like `mean`, `min`, etc...)

Another advantage of naming arguments, is that the order doesn't matter.
This is useful when there start to be more arguments. 


## Vectors and data types

```{r, echo=FALSE, purl=TRUE}
### Vectors and data types
```

A vector is the most common and basic data structure in R, and is pretty much
the workhorse of R. It's basically just a list of values, mainly either numbers or
characters. They're special lists that you can do math with. You can assign this list of values to a variable, just like you
would for one item. For example we can create a vector of :

```{r, purl=FALSE}
one_to_five <- c(1, 2, 3, 4, 5)
one_to_five <- 1:5
```

A vector can also contain characters:

```{r, purl=FALSE}
primary_colors <- c("red", "yellow", "blue")
primary_colors
```

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r, purl=FALSE}
length(one_to_five)
length(primary_colors)
```


You can also do math with whole vectors. For instance if we wanted to multiply all the values in our vector by a scalar, we can do

```{r, purl=FALSE}
5 * one_to_five
```

or we can add the data in the two vectors together

```{r, purl=FALSE}
two_to_ten <- one_to_five + one_to_five
two_to_ten
```

This is very useful if we have data in different vectors that we 
want to combine or work with. 

There are few ways to figure out what's going on in a vector. 

`class()` indicates the class (the type of element) of an object:

```{r, purl=FALSE}
class(one_to_five)
class(primary_colors)
```

The function `str()` provides an overview of the object and the elements it
contains. It is a really useful function when working with large and complex
objects:

```{r, purl=FALSE}
str(one_to_five)
str(primary_colors)
```

You can add elements to your vector simply by using the `c()` function, which stands for combine:

```{r, purl=FALSE}
new_digits <- c(one_to_five, 90) # adding at the end
new_digits <- c(30, new_digits) # adding at the beginning
new_digits
```

What happens here is that we take the original vector `one_to_five`, and we are
adding another item first to the end of the other ones, and then another item at
the beginning. We can do this over and over again to build a vector or a
dataset. As we program, this may be useful to autoupdate results that we are
collecting or calculating.

We just saw 2 of the **data types** that R uses: `"character"` and
`"integer"`. The others you will likely encounter during data analysis are:       

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"numeric"` for floating point decimal numbers      
* `"factor"` for categorical data. Similar to `"character"` data, but factors have levels   
   

Importantly, a vector can only contain **one** data type. If you combine multiple data types
in a vector with the `c()` command, R will try to coerce all the values to the same data type.
If it cannot, it will throw an error. 
    
For example, what data type is our `one_to_five` vector if we divide it by 2?
```{r, purl=FALSE}
divided_integers <- one_to_five/2
divided_integers

class(divided_integers)
```

Vectors are indexed sets, which means that every value can be referred to by its order in the vector. 
R indexes start at 1. Programming languages like Fortran, MATLAB, and R start
counting at 1, because that's what human beings typically do. Languages in the C
family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.

We can index a vector in many different ways. We can specify a position of a single value, a range of values, or a vector of values. We can even specify which values to remove by their indices. 
```{r, purl=FALSE}
one_to_five[3]
one_to_five[1:3]
one_to_five[c(1,3,5)]
one_to_five[-2]
```

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), and data frames (`data.frame`) 




