---
title: Data frames
---


------------

> ## Learning Objectives
>
> * load external data (CSV files) in memory 
> * understand the concept of a `data.frame`
> * know how to access any element of a `data.frame`
> * understand what are factors and how to manipulate them

------------

# Setup
The file required for this lesson can be downlaoded by clicking on this [link](./data/cats.csv)

- Move the downloaded file into the directory for this workshop

To view your current working directory use the `getwd()` command

```{r, eval=FALSE, echo=FALSE}
getwd()
```

To set you working directory, use the `setwd()` command. We want to set the working directory to the same location as the file. For example: 
```{r, eval=FALSE}
setwd("~/Workshops/AARUG")
```




# Looking at data


You are now ready to load the data. We are going to use the R function  `read.csv()` to load the data file into memory (as a `data.frame`):

```{r, eval=TRUE,  purl=FALSE}
cats <- read.csv(file = 'data/cats.csv')
```

This statement doesn't produce any output because assignment doesn't display
anything. If we want to check that our data has been loaded, we can print the
variable's value: `cats`

```{r, eval=FALSE, purl=FALSE}
head(cats)
View(cats)
```

We've just done two very useful things.
1. We've read our data in to R, so now we can work with it in R
2. We've created a data frame (with the read.csv command) the 
standard way R works with data. 

# What are data frames?

`data.frame` is the _de facto_ data structure for most tabular data and what we
use for statistics and plotting.

A `data.frame` is a collection of vectors of identical lengths. Each vector
represents a column, and each vector can be of a different data type (e.g.,
characters, integers, factors). The `str()` function is useful to inspect the
data types of the columns.

A `data.frame` can be created by the functions `read.csv()` or `read.table()`, in
other words, when importing spreadsheets from your hard drive (or the web).

By default, `data.frame` converts (= coerces) columns that contain characters
(i.e., text) into the `factor` data type. Depending on what you want to do with
the data, you may want to keep these columns as `character`. To do so,
`read.csv()` and `read.table()` have an argument called `stringsAsFactors` which
can be set to `FALSE`:

Let's now check the __str__ucture of this `data.frame` in more details with the
function `str()`:

```{r, purl=FALSE}
str(cats)
```

# Inspecting `data.frame` objects

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a `data.frame`. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data.

* Size:
    * `dim()` - returns a vector with the number of rows in the first element, and
    the number of columns as the second element (the __dim__ensions of the object)
    * `nrow()` - returns the number of rows
    * `ncol()` - returns the number of columns
* Content:
    * `head()` - shows the first 6 rows
    * `tail()` - shows the last 6 rows
* Names:
    * `names()` - returns the column names (synonym of `colnames()` for `data.frame`
	objects)
   * `rownames()` - returns the row names
* Summary:
   * `str()` - structure of the object and information about the class, length and
	content of  each column
   * `summary()` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.



## Indexing `data.frame` objects


Our cats data frame has rows and columns (it has 2 dimensions), if we want to
extract some specific data from it, we need to specify the "coordinates" we want
from it. Row numbers come first, followed by column numbers (i.e. [row, column]).

```{r, purl=FALSE, eval=FALSE}
cats[1, 2]   # first element in the 2nd column of the data frame
cats[1, 6]   # first element in the 6th column
cats[1:3, 7] # first three elements in the 7th column
cats[3, ]    # the 3rd element for all columns
cats[, 7]    # the entire 7th column
head_meta <- cats[1:6, ] # Row 1-6 which is the same as head
```


For larger datasets, it can be tricky to remember the column number that
corresponds to a particular variable. (Are species names in column 5 or 7? oh,
right... they are in column 6). In some cases, in which column the variable will
be can change if the script you are using adds or removes columns. It's
therefore often better to use column names to refer to a particular variable,
and it makes your code easier to read and your intentions clearer.

You can do operations on a particular column, by selecting it using the `$`
sign. In this case, the entire column is a vector. You can use
`names(cats)` or `colnames(cats)` to remind yourself of the column names.
For instance, to extract all the cats' weight information from our dataset:

```{r}
cats$weight
```

In some cases, you may way to select more than one column. You can do this using
the square brackets. Suppose we wanted weight and coat information:

```{r}
cats[, c("weight", "coat")]
```

You can even access columns by column name _and_ select specific rows of interest. For example, if we wanted the weight and coat of just rows
4 through 7, we could do:

```{r}
cats[4:7, c("weight", "coat")]
```

We can can also use logical statements to select and filter items from a `data.frame`. For example, to select all rows with black cats we could use the following statement

```{r}
cats[cats$coat == "black", ]

```

let's break this down a bit. The logical statement in the brackets returns a vector of `TRUE` and `FALSE` values. 

```{r}
cats$coat == "black"
```

These `booleans` allow us to select which records we want from our `data.frame`

Another way to do this is with the function `which()`. `which()` finds the indexes of records meeting a logical statement
```{r}
which(cats$coat == "black")

```

So, we could also write 
```{r}
cats[which(cats$coat == "black"), ]
```
But that's getting really long and ugly. R is already considered somewhat of an ugly duckling among programming languages, so no reason to play into the stereotype.

We can combine logical statements and index statements
```{r}
cats[cats$coat == "black", c("coat", "weight")]

```

Finally, we can use `&`, the symbol for "and", and `|`, the symbol for "or", to make logical statements.

```{r}
cats[cats$coat == "black" & cats$likes_string == TRUE, ]

```
This statement selects all records with black cats that also like string

## Factors

Factors are used to represent categorical data. Factors can be ordered or
unordered and are an important class for statistical analysis and for plotting.

Factors are stored as integers, and have labels associated with these unique
integers. While factors look (and often behave) like character vectors, they are
actually integers under the hood, and you need to be careful when treating them
like strings.

In the data frame we just imported, let's do 
```{r, purl=TRUE}
str(cats)
```

We can see the names of the multiple columns. And, we see that 
coat is a `Factor w/ 5 levels`

When we read in a file, any column that contains text is automatically
assumed to be a factor. Once created, factors can only contain a pre-defined set values, known as
*levels*. By default, R always sorts *levels* in alphabetical order. 


You can check this by using the function `levels()`, and check the
number of levels using `nlevels()`:

```{r, purl=FALSE, eval=FALSE}
levels(cats$coat)
nlevels(cats$coat)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high") or
it is required by particular type of analysis. Additionally, specifying the
order of the levels allows to compare levels:

```{r, purl=FALSE, error=TRUE, eval=FALSE}
satisfaction <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(satisfaction)
satisfaction <- factor(satisfaction, levels = c("low", "medium", "high"))
levels(satisfaction)
min(satisfaction) ## doesn't work
satisfaction <- factor(satisfaction, levels = c("low", "medium", "high"), ordered = TRUE)
levels(satisfaction)
min(satisfaction) ## works!
```

In R's memory, these factors are represented by numbers (1, 2, 3). They are
better than using simple integer labels because factors are self describing:
`"low"`, `"medium"`, and `"high"`" is more descriptive than `1`, `2`, `3`. Which
is low?  You wouldn't be able to tell with just integer data. Factors have this
information built in. It is particularly helpful when there are many levels
(like the species in our example data set).

### Converting factors

If you need to convert a factor to a character vector, simply use
`as.character(x)`.

Converting a factor to a numeric vector is however a little trickier, and you
have to go via a character vector. Compare:

```{r, purl=TRUE, eval=FALSE}
f <- factor(c(1, 5, 10, 2))
as.numeric(f)               ## wrong! and there is no warning...
as.numeric(as.character(f)) ## works...
as.numeric(levels(f))[f]    ## The recommended way.
```

