---
layout: topic
title: Data Visualization
author: Data Carpentry contributors
minutes: 15
---

```{r, echo=FALSE}
metadata <- read.csv('./data/Ecoli_metadata.csv')
```

> ## Learning Objectives {.objectives}
>
> * Basic plots 
> * Advanced plots (introducing `ggplot`)
> * Writing images (and other things) to file

# Basic plots in R

The mathematician Richard Hamming once said, "The purpose of computing is insight, not numbers", and the best way to develop insight is often to visualize data. Visualization deserves an entire lecture (or course) of its own, but we can explore a few features of R's plotting packages.

When we are working with large sets of numbers it can be useful to display that information graphically. R has a number of built-in tools for basic graph types such as hisotgrams, scatter plots, bar charts, boxplots and much [more](http://www.statmethods.net/graphs/). We'll test a few of these out here on the `genome_size` vector from our metadata.


```{r simplestats}
genome_size <- metadata$genome_size

```

## Scatterplot
Let's start with a **scatterplot**. A scatter plot provides a graphical view of the relationship between two sets of numbers. We don't have a variable in our metadata that is a continous variable, so there is nothing to plot it against but we can plot the values against their index values just to demonstrate the function.

```{r scatter-plot1, fig.align='center'}
plot(genome_size)
```

Each point represents a clone and the value on the x-axis is the clone index in the file, where the values on the y-axis correspond to the genome size for the clone. For any plot you can customize many features of your graphs (fonts, colors, axes, titles) through [graphic options](http://www.statmethods.net/advgraphs/parameters.html)
For example, we can change the shape of the data point using `pch`.

```{r, fig.align='center'}
plot(genome_size, pch=8)
```

We can add a title to the plot by assigning a string to `main`:

```{r, fig.align='center'}
plot(genome_size, pch=8, main="Scatter plot of genome sizes")
```

## Histogram
Another way to visualize the distribution of genome sizes is to use a histogram, we can do this buy using the `hist` function:

```{r}
hist(genome_size)
```

##Boxplot

Using additional information from our metadata, we can use plots to compare values between the different citrate mutant status using a **boxplot**. A boxplot provides a graphical view of the median, quartiles, maximum, and minimum of a data set. 

```{r boxplot, fig.align='center'}
# Boxplot
boxplot(genome_size ~ cit, metadata)
```

Similar to the scatterplots above, we can pass in arguments to add in extras like plot title, axis labels and colors.

```{r, fig.align='center'}
boxplot(genome_size ~ cit, metadata,  col=c("pink","purple", "darkgrey"),
        main="Average expression differences between celltypes", ylab="Expression")
```


# Advanced figures (`ggplot2`)

More recently, R users have moved away from base graphic options and towards a plotting package called [`ggplot2`](http://docs.ggplot2.org/) that adds a lot of functionality to the basic plots seen above. The syntax takes some getting used to but it's extremely powerful and flexible. We can start by re-creating some of the above plots but using ggplot functions to get a feel for the syntax.

`ggplot` is best used on data in the `data.frame` form, so we will will work with `metadata` for the following figures. Let's start by loading the `ggplot2` library.

```{r}
library(ggplot2)
```

The `ggplot()` function is used to initialize the basic graph structure, then we add to it. The basic idea is that you specify different parts of the plot, and add them together using the `+` operator.

We will start with a blank plot and will find that you will get an error, because you need to add layers.

```{r, eval=FALSE}
ggplot(metadata) # note the error 
```

Geometric objects are the actual marks we put on a plot. Examples include:

* points (`geom_point`, for scatter plots, dot plots, etc)
* lines (`geom_line`, for time series, trend lines, etc)
* boxplot (`geom_boxplot`, for, well, boxplots!)

A plot must have at least one geom; there is no upper limit. You can add a geom to a plot using the + operator

```{r, eval=FALSE}
ggplot(metadata) +
  geom_point() # note what happens here
```

Each type of geom usually has a required set of aesthetics to be set, and usually accepts only a subset of all aesthetics --refer to the geom help pages to see what mappings each geom accepts. Aesthetic mappings are set with the aes() function. Examples include:

* position (i.e., on the x and y axes)
* color ("outside" color)
* fill ("inside" color) shape (of points)
* linetype
* size

To start, we will add position for the x- and y-axis since `geom_point` requires mappings for x and y, all others are optional.

```{r}
ggplot(metadata) +
  geom_point(aes(x = sample, y= genome_size))
```

The labels on the x-axis are quite hard to read. To do this we need to use the themes. The ggplot2 `theme` system handles non-data plot elements such as:

* Axis labels
* Plot background
* Facet label backround
* Legend appearance

There are built-in themes we can use, or we can adjust specific elements. For our figure we will change the x-axis labels to be plotted on a 45 degree angle with a small horizontal shift to avoid overlap. 

```{r}
ggplot(metadata) +
  geom_point(aes(x = sample, y= genome_size)) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```



Resources:
---------
* [RStudio cheatsheet](http://www.rstudio.com/resources/cheatsheets/)
* [Choosing a good chart] (http://extremepresentation.typepad.com/blog/2006/09/choosing_a_good.html)
* [Nature Points of Significance] (http://www.nature.com/collections/qghhqm/pointsofsignificance)
* [Kick the bar chart habit] (http://www.nature.com/nmeth/journal/v11/n2/pdf/nmeth.2837.pdf)
